---
title: 'Exploring Speed Dating'
author: 'Colin LEVERGER'
date: '22 Oct 2016'
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: yeti
    highlight: tango
---
# Introduction

I am going to explore a Speed Dating dataset provided by Kaggle, which can be found at the following URL: https://www.kaggle.com/annavictoria/speed-dating-experiment/.

```{r, include=FALSE}
setwd("/Users/colinleverger/Documents/Experiments/kaggle/speed-dating-experiment")
```

# Initialisations
## Packages

Let's first load a few libraries...

```{r, results="hide", warning=FALSE, message=FALSE}
# Load libraries & packages
library(dplyr)        # Data manipulation
library(reshape2)     # Data reshaping for ggplot
library(ggplot2)      # Data visualization
library(RColorBrewer) # Colors on plots
library(readr)        # CSV file I/O, e.g. the read_csv function
library(dataQualityR) # DQR generation
```

## Color palettes for plots

This is optional, but if we decide to change the color of the `ggplot` afterwards, it could be useful.

```{r}
cbPalette  <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

## Read the data
```{r, results="hide", warning=FALSE, message=FALSE}
missing.types <- c("NA", "")
df <- read.csv("input/Speed Dating Data.csv", na.strings = missing.types, stringsAsFactors = F)
# "income", "tuition" and "mn_sat" features should be read as numerical
df$income  <- as.numeric(gsub(",", "", df$income))
df$tuition <- as.numeric(gsub(",", "", df$tuition))
df$mn_sat  <- as.numeric(gsub(",", "", df$mn_sat))
```

# Investigations & cleaning of the data

## Generate the Data Quality Report, before the feature engineering

The Data Quality Report (DQR) is a good way to have an overall view of the quality of your dataset. Which feature has the most missing values? How many unique values are present? Etc. It is a very good help to understand and clean the data.

```{r}
checkDataQuality(
  df,
  out.file.num = "DQR_cont-before.csv",
  out.file.cat = "DQR_cat-before.csv"
)

# Read the DQR on the disk
dqr.cont.bef <- read.csv("DQR_cont-before.csv")
dqr.cat.bef  <- read.csv("DQR_cat-before.csv")

# Classify features in function of their missingness for better analysis
feat.cont.missing20_100 <- dqr.cont.bef[dqr.cont.bef$missing.percent > 20,]
feat.cont.missing20     <- dqr.cont.bef[dqr.cont.bef$missing.percent <= 20 & dqr.cont.bef$missing.percent > 0,]
```

## Why is there so much 79 missing values? 

If we take a closer look at the data, we notice that there are a lot of features which have exactly 79 missing values.

```{r}
head(dqr.cont.bef[dqr.cont.bef$missing == 79, c(1,2,3)], 10)
```

The question is: is there a pattern? Can we impute, how to clean this mess?

```{r}
val1 <- feat.cont.missing20[feat.cont.missing20$missing == 79,]$X
head(df[is.na(df$imprelig), c(1:10)])
```

It appears that nothing very interesting can be deducted from this.

## Investigate with missing age_o/race_o

In the same approach, there is a lot of missing values for `age_o` and `race_o` features.

```{r, results="hide", warning=FALSE, message=FALSE}
missing.age_o.pid <- unique(df[is.na(df$age_o),]$pid)
missing.race_o.pid <- unique(df[is.na(df$race_o),]$pid)
head(subset(df, iid %in% missing.age_o.pid)[, c(1:10)])
# Nothing very interesting here neither...
```

## Deal with missing id

According to our DQR, there is one missing `id` in our dataset. It should be fairly easy to deduct the good value. Let's find and replace it.

```{r}
# Find the iid of the only missing id
iid <- df[is.na(df$id),]$iid
# Assign this iid' id to the missing iid..
df[is.na(df$id),]$id <- head(df[df$iid == iid,]$id, 1)
```

_Note to myself: it is not useful to make the process generic, the data won't change anyway..._
 
## Deal with missing pid

There is 10 missing `pid` (partner’s iid number) in the dataset. Since there is no `iid` missing, we could probably input this last easily. We could see if the `partner` value is present, because it will lead us within the wave to this missing `pid`.

```{r}
# Show the missing pid
df[is.na(df$pid), c(1,2,11,12)]
# Save the partner number for the wave
partner.pid <- unique(df[is.na(df$pid),]$partner)
# Save the wave number
wave.pid <- unique(df[is.na(df$pid),]$wave)
# Show the iid we are looking for
unique(df[df$wave == wave.pid & df$id == partner.pid,]$iid)
df[is.na(df$pid),]$pid <- 128
```

## Work on field feature
```{r}
# Plot raw field
df$field <- tolower(df$field)
barplot(
  table(df$field),
  main = "Careers"
)

# How many unique field do we have?
length(unique(df$field))

# Show coded field
ordered(unique(df$field_cd))
```

## Work on career feature
```{r}
# Plot raw career
df$career <- tolower(df$career)
barplot(
  table(df$career),
  main = "Careers"
)
# How many unique carrer do we have?
length(unique(df$career))

# Show coded career
ordered(unique(df$career_c))
```

## Work on zip codes
```{r}
# Assign NA to all the zip codes equals to 0
df[df$zipcode == 0 & !is.na(df$zipcode),]$zipcode <- NA
```

## Standardise wave > 5 and < 10

It is said on the word doc linked to the data that the waves 6 to 9 are different because people were asked to note their preferences from 1 to 10 rather than allocating a hundred points on features.

```{r}
# Show waves 6 to 9
head(df[df$wave > 5 & df$wave < 10, c(1,16:20)])
# Show wave 3 to compare
head(df[df$wave == 3, c(1,16:20)])
```

However, we see above that the data seems to already be standardised - no work for us here!

## Change Male and Female attributes

To enhance comprehension, I have chosen to display `W` instead if `0` for women, and `M` instead of `1` for males.

```{r}
df[df$gender == 0,]$gender <- "W"
df[df$gender == 1,]$gender <- "M"
```

## Generate DQR, after feature engineering
```{r}
checkDataQuality(
  df,
  out.file.num = "DQR_cont-after.csv",
  out.file.cat = "DQR_cat-after.csv"
)
```

## Read DQR on the disk & check that everything is fine
```{r, results="hide"}
dqr.cont.aft <- read.csv("DQR_cont-after.csv")
dqr.cat.aft  <- read.csv("DQR_cat-after.csv")

dqr.cont.aft
dqr.cat.aft
```

# Analyse the data
## Gender analysis
```{r, warning=FALSE, message=FALSE}
gender.rep.over.waves <- subset(df, !duplicated(df[, 1])) %>%
  group_by(wave, gender) %>%
  summarise(my.n = n()) %>%
  melt(id.vars = c("gender", "wave"))

# Plot gender repartition in waves
ggplot(gender.rep.over.waves, aes(x = wave, y = value, fill = factor(gender))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_discrete(name = "Gender") +
  xlab("Wave") + ylab("Population") + 
  scale_fill_manual(values = cbPalette)
```

## Age analysis
```{r, warning=FALSE, message=FALSE}
age.rep.over.waves <- subset(df, !duplicated(df[, 1])) %>%
  filter(!is.na(age)) %>%
  group_by(wave, gender) %>%
  summarise(my.m = mean(age)) %>%
  melt(id.vars = c("gender", "wave"))

# Plot age repartition in waves
ggplot(age.rep.over.waves, aes(x = wave, y = value, fill = factor(gender))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_discrete(name = "Gender") +
  xlab("Wave") + ylab("Mean age") +
  scale_fill_manual(values = cbPalette)

# What are the extremums?
print(paste0("Minimum age in all the dataset is: ", min(df[!is.na(df$age),]$age)))
print(paste0("Max age in all the dataset is: ", max(df[!is.na(df$age),]$age)))
print(paste0("Mean age in all the dataset is: ", mean(df[!is.na(df$age),]$age)))
```

## Work and studies analysis
```{r}
# Create study field codes
fields.cd <- c(
  "Law",
  "Math",
  "Social Science, Psychologist" ,
  "Medical Science, Pharmaceuticals, and Bio Tech",
  "Engineering",
  "English/Creative Writing/ Journalism",
  "History/Religion/Philosophy",
  "Business/Econ/Finance",
  "Education, Academia",
  "Biological Sciences/Chemistry/Physics",
  "Social Work" ,
  "Undergrad/undecided" ,
  "Political Science/International Affairs" ,
  "Film",
  "Fine Arts/Arts Administration",
  "Languages",
  "Architecture",
  "Other"
)

# Create career codes
career.cd <- c(
  "Lawyer",
  "Academic/Research", 
  "Psychologist", 
  "Doctor/Medicine",
  "Engineer", 
  "Creative Arts/Entertainment",
  "BankingBusiness/CEO/Admin",
  "Real Estate",
  "International/Humanitarian Affairs",
  "Undecided" ,
  "Social Work",
  "Speech Pathology",
  "Politics",
  "Pro sports/Athletics",
  "Other",
  "Journalism",
  "Architecture"
)

# Find number of men/women on each study field
fields <- df[!is.na(df$field_cd),] %>%
  group_by(gender, field_cd) %>%
  summarise(
    my.n = n()
  )

# Find number of men/women on each career
careers <- df[!is.na(df$career_c),] %>%
  group_by(gender, career_c) %>%
  summarise(
    my.n = n()
  )

# Plot study fields repartition
ggplot(fields, aes(x = field_cd, y = my.n, fill = factor(gender))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_discrete(name = "Gender") +
  xlab("Field") + ylab("Count") +
  scale_x_continuous(labels = fields.cd, breaks = 1:18) +
  coord_flip()

# Plot careers repartition
ggplot(careers, aes(x = career_c, y = my.n, fill = factor(gender))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_discrete(name = "Gender") +
  xlab("Career") + ylab("Count") +
  scale_x_continuous(labels = career.cd, breaks = 1:17) +
  coord_flip()
```

## Race analysis
```{r}
# Create race code
race.c <- c(
  "Black/African American",
  "European/Caucasian-American",
  "Latino/Hispanic American",
  "Asian/Pacific Islander/Asian-American",
  "Native American",
  "Other"
)

races <- df[!is.na(df$race),] %>%
  group_by(gender, race) %>%
  summarise(
    my.n = n()
  )

# Plot race repartition
ggplot(races, aes(x = race, y = my.n, fill = factor(gender))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_discrete(name = "Gender") +
  xlab("Race") + ylab("Count") +
  scale_x_continuous(labels = race.c, breaks = 1:6) +
  coord_flip()
```

## Goal analysis
```{r}
# Create goal code
goal.c <- c(
  "Seemed like a fun night out",
  "To meet new people",
  "To get a date",
  "Looking for a serious relationship",
  "To say I did it",
  "Other"
)

goals <- df[!is.na(df$goal),] %>%
  group_by(gender, goal) %>%
  summarise(
    my.n = n()
  )

# Plot goals repartition
ggplot(goals, aes(x = goal, y = my.n, fill = factor(gender))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_discrete(name = "Gender") +
  xlab("Goal") + ylab("Count") +
  scale_x_continuous(labels = goal.c, breaks = 1:6) +
  coord_flip()
```

## Date & go out analysis
```{r}
# Create date & go out code
date.c <- c(
  "Several times a week",
  "Twice a week",
  "Once a week",
  "Twice a month",
  "Once a month",
  "Several times a year",
  "Almost never"
)

dates <- df[!is.na(df$date),] %>%
  group_by(gender, date) %>%
  summarise(
    my.n = n()
  )

go.outs <- df[!is.na(df$go_out),] %>%
  group_by(gender, go_out) %>%
  summarise(
    my.n = n()
  )

# Plot dates repartition
ggplot(dates, aes(x = date, y = my.n, fill = factor(gender))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_discrete(name = "Gender") +
  xlab("Date") + ylab("Count") +
  scale_x_continuous(labels = date.c, breaks = 1:7) +
  coord_flip()

# Plot go out repartition
ggplot(go.outs, aes(x = go_out, y = my.n, fill = factor(gender))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_discrete(name = "Gender") +
  xlab("Go out") + ylab("Count") +
  scale_x_continuous(labels = date.c, breaks = 1:7) +
  coord_flip()
```

## Matches analysis
```{r}
# Dummie analysis on Matches
barplot(
  table(df$match),
  main = "Matches proportion",
  col = "black"
)
```

### Match by gender analysis
```{r}
match.by.gender <- df %>%
  group_by(gender) %>%
  summarise(
    nb_matches = sum(match == 1),
    nb_fails = sum(match == 0)
  ) %>% 
  melt(id.vars = "gender")

# Plot matches for both men and women
ggplot(match.by.gender, aes(x = variable, y = value, fill = factor(gender))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_discrete(name = "Gender") +
  xlab("Result") + ylab("Count")
```

### Match by wave analysis
```{r}
match.by.waves <- df[df$match == 1,] %>%
  group_by(wave) %>%
  summarise(
    nb_matches = sum(match == 1)
  )

# Plot matches for waves: what was the best wave to be?
ggplot(match.by.waves, aes(x = wave, y = nb_matches)) +
  geom_bar(stat = "identity", position = "dodge") +
  xlab("Wave number") + ylab("Matches")
```

## Look for best attributes to have for a man
```{r}
# TODO
```

# Create data to feed Gephi
```{r}
# Isolate iid (1) -> pid (12)
gephi <- df[,c(1, 12)]
# Rename columns to match Gephi's prerequisites
names(gephi)[1] <- paste("Source")
names(gephi)[2] <- paste("Target")
# Convert Target into integer (to avoid having 10.0, 11.0...)
# We don't want to have ".0" to have a good analysis with Gephi
gephi$Target <- as.integer(gephi$Target)
# Write on disk
write_csv(gephi, "gephi.csv")
```
